---
title: "BeerSales"
author: "Justin Howard, Amber Burnet"
date: "June 17, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir ='C:/Users/howar/Documents/doingdatascience/BeerSales')
```
## Loading the beer data:
```{r loading beer data}
file1<- 'C:/Users/howar/Documents/R__working_directory/DDS/msds-6306-doing-data-science/Unit 7/Beers.csv'
Beers<- read.csv(file1)
str(Beers)
head(Beers)
```
## Loading the brewery data:
```{r loading brewiery data}
file2<- 'C:/Users/howar/Documents/R__working_directory/DDS/msds-6306-doing-data-science/Unit 7/Breweries.csv'
Breweries<- read.csv(file2)
str(Breweries)
head(Breweries)
```
## Merging the two datasets on "Brewery_id"
```{r merging dataset}
names(Breweries)<- c("Brewery_id", "Brewery_Name", "City", "State")
BeerNBrew<- merge(Beers, Breweries, by = "Brewery_id")
head(BeerNBrew)
tail(BeerNBrew)
```
## Counting the number of NA values per column:
```{r NA counts}
na_count<- sapply(BeerNBrew, function(x) sum(is.na(x)))
na_count<- data.frame(na_count)
na_count
```  
#### Observations:
2 columns contain large numbers of NA values: 
ABV - 62
IBU - 1005

## How many breweries are in each state?
```{r aggregating, warning=F}
library(dplyr)
bystate<- BeerNBrew %>%
  group_by(State) %>%
  distinct(Brewery_id)
  
head(bystate)
brewcount<- aggregate(Brewery_id ~ State, BeerNBrew, length)

names(brewcount)<- c("State", "Brewery_Count")
head(brewcount)
```
## Visualize data with a US map
First, prepare the data for the map by adding in state names.
```{r}
#trimming out whitespace
brewcount$State<- trimws(brewcount$State, which = "both")
# a csv with state names matched to abbreviations
# file<- download.file('http://www.whypad.com/wp-content/uploads/us_states.zip')
state.abbv<- read.csv('us_states.csv', header = F)
#give the columns names to reference by
names(state.abbv)<- c("State.Caps", "State.Name", "State")
#select only data we need
addin<- state.abbv %>%
  select(State.Name, State)
# make all state names lowercase to match states data
addin$State.Name<- tolower(addin$State.Name)
#merge with brewcount dataframe
brewcount.states<- merge(brewcount, addin, by = "State")

#loading packages
library(ggmap)
library(maps)
library(mapdata)

#make dataframe of states and merge with our data
states <- map_data("state")
addbrews<- brewcount.states %>%
  select(Brewery_Count, State.Name)
names(addbrews)<- c("Brewery_Count", "region")
#brewery.map.data<-merge(states, addbrews, by = "region", all.x = T)
#using merge in this case causes rips in the map, instead, use left_join from dplyr
brewery.map.data2<- left_join(states, addbrews)
```
Make a map with colors based on the number of breweries in each state. We will perform a log transformation on the counts to create a more even standard deviation, which will improve the visualization by making differences more evident.
```{r}
ggplot(data = brewery.map.data2) + 
  ggtitle("Breweries Per State") +
  theme(plot.title = element_text(hjust = .5)) +
  geom_polygon(aes(x = long, y = lat, fill = Brewery_Count, group = group), color = "white") + 
  coord_fixed(1.3) + theme_nothing(legend = T) + guides(fill=guide_legend(title = "Number of Breweries"))
```

Colorado is clearly an outlier for the number of breweries it hosts. This is helpful, but we will also examine breweries per capita by adding state populations to our dataset. Our data come from the US Census Bureau.

```{r}

#load data with state populations
pops<- read.csv('C:/Users/howar/Documents/doingdatascience/beersales/statepopulations.csv', head=T)
pops<- pops %>%
  select(NAME, POPESTIMATE2018)
pops$NAME<- tolower(pops$NAME)
names(pops) <- c("State.Name", "PopEst2018")
head(pops)

#merging with state names table
brewcount.states<- left_join(brewcount.states, pops)

brewcount.states<- brewcount.states %>%
  mutate(Brews.Per.Capita = Brewery_Count / PopEst2018)

# making the new map table
add.per.capita<- brewcount.states %>%
  select(Brews.Per.Capita, State.Name)
names(add.per.capita)<- c("Brews.Per.Capita", "region")

brews.per.capita.map<- left_join(states,add.per.capita)
brews.per.capita.map$Brews.Per.Capita<- round(brews.per.capita.map$Brews.Per.Capita, digits = 6)*100000

ggplot(data = brews.per.capita.map) + 
  ggtitle("Breweries Per Capita") +
  theme(plot.title = element_text(hjust = .5)) +
  geom_polygon(aes(x = long, y = lat, fill = Brews.Per.Capita, group = group), color = "white") + 
  coord_fixed(1.3) + theme_nothing(legend = T) + guides(fill=guide_legend(title = "Breweries per 100,000 People")) 
```

We see that Colorado is still an outlier when we look at the number of breweries per person. Other notable outliers include Vermont, Montana, Oregon and Wyoming. 


## Median alcohol content and international bitterness unit for each state.
```{r aggregating medians - IBU}
IBUvals<- BeerNBrew %>%
  group_by(State) %>%
  filter(!is.na(IBU))
head(IBUvals)

IBUmedians<- aggregate(IBU ~ State, IBUvals, median)
names(IBUmedians)<- c("State", "Median_IBU")
head(IBUmedians)
```

```{r aggregating medians - alcohol content}
boozecontent<- BeerNBrew %>%
  group_by(State)
boozemedians<- aggregate(ABV ~ State, boozecontent, median)
names(boozemedians)<- c("State", "Median_ABV")
boozemedians<- arrange(boozemedians, boozemedians$Median_ABV)
```
### Bar Charts 
```{r plotting the medians}
library(ggplot2)
ggplot(IBUmedians, aes(x=reorder(State, -Median_IBU), y=Median_IBU, fill=State)) + geom_bar(stat= 'identity') + guides(fill=F)+ theme(plot.title = element_text(hjust= .5), axis.text.x = element_text(angle=90, vjust = .5)) +
  ggtitle("Median Bitterness by State") + xlab('State') + ylab('Median IBU Value')
ggplot(boozemedians, aes(x=reorder(State, -Median_ABV), y=(Median_ABV-.039), fill=Median_ABV)) + geom_bar(stat= 'identity') + 
  guides(fill=F) + theme(plot.title = element_text(hjust= .5), axis.text.x = element_text(angle=90, vjust = .5)) +
  ggtitle("Median Alcohol Volumes by State") + ylab('Median - .039')
```

Prepping and merging the median IBU the data with the states data. 
```{r prepping and merging IBU map}
#trimming out whitespace
IBUmedians$State<- trimws(IBUmedians$State, which = "both")

#merge with IBUmedians dataframe
IBUmedians.states<- merge(IBUmedians, addin, by = "State")

addIBU<- IBUmedians.states %>%
  select(Median_IBU, State.Name)
names(addIBU)<- c("Median_IBU", "region")

IBU.map<- left_join(states, addIBU)

ggplot(data = IBU.map) + 
  ggtitle("Median Bitterness by State (IBU Units)") +
  theme(plot.title = element_text(hjust = .5)) +
  scale_fill_gradientn(colours=terrain.colors(10)) +
  geom_polygon(aes(x = long, y = lat, fill = Median_IBU, group = group), color = "white") + 
  coord_fixed(1.3) + theme_nothing(legend = T) + guides(fill=guide_legend(title = "Bitterness Scale"))
```

The map indicates that Midwestern states tend to contain more breweries with low median reported bitterness than other regions. West Coast states appear to consistently produce beers in the medium bittnerness range, while bitterness varies widely on the East Coast. The East Coast does contain 3 of the states with the highest median bitterness

The wide variance among states indicates that breweries that have reported their bitterness ratings may have a sginficant influence on the available data. Some breweries that specialize in bitter beers make a significant impact on the dataset and could be significantly impacting the ptaste preferences of their entire state.

We noticed that North Dakota is grey and foiund that the dataset contained no bittnerness data on North Dakota. No bittnerness ratings were available from breweries in that state.

```{r prepping and merging for ABV map}
#trimming out whitespace
boozemedians$State<- trimws(boozemedians$State, which = "both")

#merge with IBUmedians dataframe
boozemedians.states<- merge(boozemedians, addin, by = "State")

addABV<- boozemedians.states %>%
  select(Median_ABV, State.Name)
names(addABV)<- c("Median_ABV", "region")

ABV.map<- left_join(states, addABV)

ggplot(data = ABV.map) + 
  ggtitle("Median ABV by State") +
  scale_fill_gradientn(colours=terrain.colors(10))+
  theme(plot.title = element_text(hjust = .5)) +
  geom_polygon(aes(x = long, y = lat, fill = Median_ABV, group = group), color = "white") + 
  coord_fixed(1.3) + theme_nothing(legend = T) + guides(fill=guide_legend(title = "ABV Scale"))
```

Utah is a clear outlier, with a near 0 ABV, followed by New Jersey...of all places! Among the states that have the highest median alcohol content are Kentucky and West Virginia. Although it is too small to be shown, our nation's capital allows the highest alcohol content by volume, topping  

## Maximum Alcoholic(ABV) Beer by State
```{r maximum - abv content,echo=TRUE}
 maxABV<-max(BeerNBrew$ABV,na.rm=TRUE)
statemaxABV<-filter(BeerNBrew, ABV==maxABV)
statemaxABV %>% select(State,ABV)
statemaxABV
```
## Top 3 States with Highest ABV
``` {r top3ABV, echo=TRUE}
newdata <- BeerNBrew[order(-BeerNBrew$ABV, na.last=TRUE),] 
l<-head(newdata,3)
mm<-ggplot(l, aes(x=State,y=ABV,fill=State))+geom_bar(stat="identity")+xlab("State")+ylab("ABV")+ggtitle("Top Three States with Highest ABV")+theme_minimal()
mm
```
## Questin 5 Part B: Maximum IBU Bitterness by State
```{r maximum - ibu content,echo=TRUE}
 maxIBU<-max(BeerNBrew$IBU,na.rm=TRUE)
statemaxIBU<-filter(BeerNBrew, IBU==maxIBU)
statemaxIBU %>% select(State,IBU)
```
## Top 3 States with Highes IBU
``` {r top3IBU, echo=TRUE}
newdata2 <- BeerNBrew[order(-BeerNBrew$IBU, na.last=TRUE),] 
head(newdata2,3)
n<-head(newdata2,3)
nn<-ggplot(n, aes(x=State,y=IBU,fill=State))+geom_bar(stat="identity")+xlab("State")+ylab("IBU")+ggtitle("Top Three States with Highest IBU")+theme_minimal()+scale_fill_brewer()
nn
```
## Question 6 Summary Statistics for the ABV Variable by Summary Function
``` {r summarystatABV, echo=TRUE}
dfa<-(BeerNBrew$ABV)
summary(dfa)
ggplot(BeerNBrew, aes(y=ABV, x=" ", fill= " ")) + geom_boxplot() + ggtitle("Summary of Alcohol Content") +
  theme(plot.title = element_text(hjust = .5), legend.position = "none") + ylab("Percentage of Alcohol Content") +
  xlab("ABV") + coord_flip()
```

## Question 7: ABV and IBU Correlation.  There is a positive correlatin between the two.
``` {r scatterplot, echo=TRUE}
library(ggplot2)
p1 <- ggplot(BeerNBrew, aes(IBU, ABV)) + 
  geom_point(size = 3) + geom_smooth(method='lm') + 
  labs(x = "IBU", y = "ABV") + theme_bw() +
  theme(legend.position = "top", legend.text = element_text(size = 12), 
        legend.title = element_text(size = 14))+ggtitle("Correlation between Bitterness and Alcohol Content")
p1
```